\documentclass[12pt, a4paper]{article}
\input{../preamble}

\begin{document}

\begin{center}
  {\LARGE \textbf{Chapter 01: The Euler Method}}\\[0.5em]
  {\large \textcolor{hintgray}{Ordinary Differential Equations — From Taylor Series to Numerical Integration}}\\[0.3em]
  {\small \textcolor{hintgray}{Rocket Science Workbook — simulation-2D-jl}}
\end{center}
\vspace{1em}\hrule\vspace{1.5em}

% ──────────────────────────────────────────────────────────────
\section{What is an Ordinary Differential Equation?}

\begin{theorybox}{Definition}
An \textbf{ordinary differential equation (ODE)} relates an unknown function $y(t)$ to its own derivatives:
\[
  \frac{dy}{dt} = f(t,\, y), \qquad y(t_0) = y_0
\]
The pair of the ODE and the initial condition $y(t_0)=y_0$ is called an \textbf{initial value problem (IVP)}.
The goal: find $y(t)$ for all $t > t_0$.
\end{theorybox}

\subsection*{Analytical Solution via Separation of Variables}

Many first-order ODEs can be solved exactly. The technique of \textbf{separation of variables} rewrites the ODE so that all $y$-terms are on one side and all $t$-terms on the other, then integrates both sides.

\textbf{Example:} $\dfrac{dy}{dt} = -k\,y$

\begin{align*}
  \frac{dy}{y} &= -k\,dt                   && \text{(separate)} \\
  \int \frac{1}{y}\,dy &= \int -k\,dt      && \text{(integrate both sides)} \\
  \ln|y| &= -kt + C                         && \text{(antiderivatives)} \\
  y(t) &= y_0\,e^{-kt}                      && \text{(exponentiate, apply IC)}
\end{align*}

This is the \textbf{exact} solution. Any numerical method must approximate this.

% ──────────────────────────────────────────────────────────────
\section{The Taylor Series — Foundation of All Numerical Methods}

\begin{theorybox}{Taylor Expansion of $y(t+h)$}
If $y(t)$ is smooth, we can expand it around $t$:
\[
  y(t+h) = y(t) + h\,\underbrace{y'(t)}_{= f(t,y)} + \frac{h^2}{2!}\,y''(t) + \frac{h^3}{3!}\,y'''(t) + \cdots
\]
This is an \textbf{exact} formula (infinite series). Every numerical method is derived by truncating this series at some order.
\end{theorybox}

\subsection*{Second Derivative via the Chain Rule}

The second derivative $y''(t)$ can be expressed using $f$:
\[
  y''(t) = \frac{d}{dt}f(t,y(t)) = \underbrace{\frac{\partial f}{\partial t}}_{\text{explicit}} + \underbrace{\frac{\partial f}{\partial y}\cdot f(t,y)}_{\text{implicit (chain rule)}}
\]
This is why higher-order Taylor methods require computing partial derivatives of $f$ — which is expensive. Runge-Kutta methods (Chapter 02) achieve the same accuracy without this.

\begin{theorybox}{The Euler Method from Taylor Truncation}
Truncate after the first derivative term:
\[
  y(t+h) \approx y(t) + h\,f(t,y(t))
\]
\[
  \boxed{y_{n+1} = y_n + h\cdot f(t_n,\,y_n)}
\]
The \textbf{local truncation error} is the dropped term: $\dfrac{h^2}{2}y''(\xi) = \mathcal{O}(h^2)$.\\
After $N = T/h$ steps, errors accumulate: \textbf{global error} $= \mathcal{O}(h)$.
\end{theorybox}

% ──────────────────────────────────────────────────────────────
\section{Geometric Interpretation}

At each point $(t_n, y_n)$, the ODE tells us the \textbf{slope} of the solution curve: $f(t_n, y_n)$.
Euler's method moves a distance $h$ along this slope — a straight-line approximation.
The error arises because the true curve is not actually straight.

\begin{center}
\begin{tikzpicture}
\begin{axis}[
  width=10cm, height=6cm,
  xlabel={$t$}, ylabel={$y$},
  xmin=0, xmax=1.2, ymin=0, ymax=1.2,
  grid=both, grid style={line width=0.3pt, draw=gray!30},
  title={Euler steps vs.\ exact solution}
]
% Exact: e^{-t}
\addplot[domain=0:1.2, samples=100, thick, black] {exp(-x)};
\addlegendentry{Exact $e^{-t}$}
% Euler h=0.4: points at 0, 0.4, 0.8, 1.2
% y0=1, y1=0.6, y2=0.36, y3=0.216
\addplot[blue, mark=*, mark size=2pt, thick] coordinates {
  (0,1) (0.4, 0.6) (0.8, 0.36) (1.2, 0.216)
};
\addlegendentry{Euler $h=0.4$}
\end{axis}
\end{tikzpicture}
\end{center}

% ──────────────────────────────────────────────────────────────
\section{Stability Analysis}

\begin{theorybox}{The Test Equation}
The canonical stability test equation is $\dot{y} = \lambda y$ with $\lambda \in \mathbb{C}$.\\
Applying Euler: $y_{n+1} = y_n + h\lambda y_n = (1 + h\lambda)\,y_n$

After $n$ steps: $y_n = (1+h\lambda)^n y_0$.\\
For stability we need $|1 + h\lambda| \leq 1$.

\textbf{Real case ($\lambda < 0$):} Stability requires $-2 \leq h\lambda \leq 0$, i.e.\
\[
  h \leq \frac{2}{|\lambda|}
\]
This is the \textbf{stability condition} — the step size has an upper bound!
\end{theorybox}

% ──────────────────────────────────────────────────────────────
\section{Exercises}

\begin{exercisebox}{1 — Analytical Solution: Separation of Variables}
Solve the following ODEs analytically using separation of variables.
Show every step of the integration.

\textbf{a)} $\dfrac{dy}{dt} = 3y$, \quad $y(0) = 2$

\vspace{0.3em}
\textit{Step 1 (separate):} $\dfrac{dy}{y} = $ \underline{\hspace{3cm}}

\textit{Step 2 (integrate both sides):} $\displaystyle\int\dfrac{dy}{y} = \int$ \underline{\hspace{3cm}}

\textit{Step 3 (solve for $y$):} $y(t) = $ \underline{\hspace{4cm}}

\vspace{1em}
\textbf{b)} $\dfrac{dy}{dt} = -2t\,y$, \quad $y(0) = 1$
\quad (Answer: $y = e^{-t^2}$)

\vspace{0.3em}
Show full derivation:
\end{exercisebox}

\begin{solutionspace}[9cm]
\end{solutionspace}

\begin{exercisebox}{2 — Taylor Series Expansion}
\textbf{a)} Write the Taylor expansion of $e^{-h}$ around $h=0$ up to order $h^4$.
\[
  e^{-h} = \sum_{n=0}^{\infty} \frac{(-h)^n}{n!} = \underline{\hspace{8cm}}
\]

\textbf{b)} The Euler approximation for $\dot{y} = -y$, $y(0)=1$ after one step of size $h$ gives $y_1 = 1-h$.
The exact value is $y(h) = e^{-h}$.
Using your expansion from (a), show that the error is:
\[
  e^{-h} - (1-h) = \underline{\hspace{6cm}} = \mathcal{O}(h^2)
\]

\textbf{c)} What is the leading error term? For $h=0.1$, estimate the error numerically.
\end{exercisebox}

\begin{solutionspace}[8cm]
\end{solutionspace}

\begin{exercisebox}{3 — Euler by Hand: Five Steps}
Given: $\dfrac{dy}{dt} = -y$, \; $y(0)=1$, \; $h=0.2$.
\quad Exact solution: $y(t) = e^{-t}$.

Fill in the table. Compute every entry by hand — use the formula
$y_{n+1} = y_n + h\cdot f(t_n, y_n) = y_n - 0.2\,y_n = 0.8\,y_n$.

\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline
$n$ & $t_n$ & $y_n$ (Euler) & $e^{-t_n}$ (exact) & $|$error$|$ \\
\hline
0 & 0.0 & 1.0000\phantom{000} & 1.0000\phantom{000} & 0.0000 \\
\hline
1 & 0.2 & & 0.8187 & \\
\hline
2 & 0.4 & & 0.6703 & \\
\hline
3 & 0.6 & & 0.5488 & \\
\hline
4 & 0.8 & & 0.4493 & \\
\hline
5 & 1.0 & & 0.3679 & \\
\hline
\end{tabular}
\end{center}

At $t=1.0$: relative error $= \dfrac{|y_5 - e^{-1}|}{e^{-1}} \approx $ \underline{\hspace{2cm}} \%
\end{exercisebox}

\begin{solutionspace}[5cm]
\end{solutionspace}

\begin{exercisebox}{4 — Error Order Verification}
Repeat Exercise 3 with $h=0.4$ (2 steps to reach $t=0.8$) and $h=0.1$ (10 steps).

\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
$h$ & $y_N$ at $t=1$ & Error $|y_N - e^{-1}|$ & Ratio (error$_{2h}$ / error$_h$) \\
\hline
0.4  & & & — \\
\hline
0.2  & (from Ex.~3) & & \\
\hline
0.1  & & & \\
\hline
\end{tabular}
\end{center}

\textbf{Expected:} Each halving of $h$ halves the error (ratio $\approx 2$). Does your data confirm $\mathcal{O}(h)$?
\end{exercisebox}

\begin{solutionspace}[6cm]
\end{solutionspace}

\begin{exercisebox}{5 — Free Fall as a Vector ODE}
A ball is dropped from $h_0 = 500\,\text{m}$ with $v_0 = 0$.
State vector $\mathbf{y} = [h,\, v]^\top$, ODE: $\dot{\mathbf{y}} = [v,\; -g]^\top$, $g=9.81\,\text{m/s}^2$.

\textbf{a)} Solve analytically: $h(t) = $ \underline{\hspace{4cm}}, \quad $v(t) = $ \underline{\hspace{4cm}}

\textbf{b)} At what time does the ball hit the ground?
\[
  h(t^*) = 0 \;\Rightarrow\; t^* = \sqrt{\frac{2h_0}{g}} = \underline{\hspace{3cm}}\,\text{s}
\]

\textbf{c)} Apply Euler with $h_\text{step} = 2\,\text{s}$. Fill in 4 steps:

\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline
$n$ & $t$ (s) & $h$ (m) & $v$ (m/s) & $h_\text{exact}$ (m) \\
\hline
0 & 0  & 500.00 & 0.00   & 500.00 \\
\hline
1 & 2  & & & \\
\hline
2 & 4  & & & \\
\hline
3 & 6  & & & \\
\hline
4 & 8  & & & \\
\hline
\end{tabular}
\end{center}

\textbf{d)} Why does Euler \textit{over}estimate the height for free fall? Sketch the argument using the Taylor series.
\end{exercisebox}

\begin{solutionspace}[8cm]
\end{solutionspace}

\begin{exercisebox}{6 — The Logistic Equation (Challenge)}
The logistic ODE models population growth with a carrying capacity $K$:
\[
  \frac{dy}{dt} = r\,y\left(1 - \frac{y}{K}\right), \quad y(0) = y_0
\]

\textbf{a)} Use \textbf{partial fractions} to separate and integrate:
\[
  \int \frac{dy}{y(1 - y/K)} = \int r\,dt
\]
\textit{Hint:} $\dfrac{1}{y(1-y/K)} = \dfrac{1}{y} + \dfrac{1/K}{1-y/K}$

\textbf{b)} Show that the exact solution is:
\[
  y(t) = \frac{K}{1 + \left(\frac{K}{y_0}-1\right)e^{-rt}}
\]

\textbf{c)} With $r=1$, $K=100$, $y_0=10$: Apply Euler with $h=0.5$ for 3 steps.
Compare to exact solution at $t=1.5$.

\textbf{d)} What happens to $y(t)$ as $t\to\infty$? Find the equilibrium analytically by setting $\dot{y}=0$.
\end{exercisebox}

\begin{solutionspace}[12cm]
\end{solutionspace}

% ──────────────────────────────────────────────────────────────
\section{Resources}

\begin{hintbox}
\textbf{Watch before starting the template:}
\begin{itemize}
  \item \textbf{3Blue1Brown} — ``Differential equations'' (visual intuition, highly recommended):\\
        \texttt{youtube.com/watch?v=p\_di4Zn4wz4} (DE overview) and the full playlist
  \item \textbf{Khan Academy} — ``Separable equations'' (step-by-step technique):\\
        \texttt{khanacademy.org/math/ap-calculus-ab/ab-diff-equations-new}
  \item \textbf{Michel van Biezen} — ``Differential Equations'' (very thorough worked examples):\\
        Search: ``Michel van Biezen differential equations'' on YouTube
\end{itemize}
\textbf{Reference:}
\begin{itemize}
  \item Paul's Online Math Notes — ODEs: \texttt{tutorial.math.lamar.edu/Classes/DE/DE.aspx}
  \item MIT 18.03 Differential Equations (free): \texttt{ocw.mit.edu/courses/18-03sc-differential-equations-fall-2011}
\end{itemize}
\end{hintbox}

\section{Connection to the Julia Template}

\begin{theorybox}{What you implement in \texttt{template.jl}}
\begin{enumerate}
  \item \texttt{euler\_step(f, t, y, h)} — one Euler step (works for scalars and vectors)
  \item \texttt{euler\_solve(f, y0, t0, t\_end, h)} — full integration loop
  \item Test case 1: $\dot{y}=-y$ — compare to exact $e^{-t}$, plot error vs.\ $h$
  \item Test case 2: Free fall vector system (Exercise 5)
  \item Test case 3: Logistic equation (Exercise 6) — does your Euler stay below $K$?
\end{enumerate}
\end{theorybox}

\end{document}
